### 2014-08-05 Studies/2012_Chicago/Report/Chicago_Fire.tex
Chicago: Edits based on reviewer comments

% Extra packages
\usepackage{xfrac}

@@

\chapter*{\centering Abstract}
\addcontentsline{toc}{chapter}{Abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2 \sfrac{1}{2} story, single family structure in Chicago, IL, that resulted in the death of a fire captain. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at a height of 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

@@

On November 2, 2012, a fire in a 2~\sfrac{1}{2} story wood frame residential structure claimed the life of a Chicago fire captain. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{fire_sum} provides a summary of the fire incident, Section~\ref{model} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{results} presents the simulation results, and Section~\ref{discussion} discusses the simulation results as they relate to firefighter safety and effectiveness.

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident was documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines are ``RIT Response'' to be one truck company, one battalion chief, one ALS ambulance, and one EMS field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

@@

\noindent Table~\ref{tab:fire_info} shows an overview of the timeline of events. A plan view of the structure and the location where the captain collapsed are shown in Fig.~\ref{fig:simp_geom}.
\begin{table}
\centering
\captionof{table}{Abridged NIOSH approximate fire event timeline~\cite{NIOSH:Bowyer}}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time  &  Fire Behavior / Fireground Operation  \\
{[hh:mm:ss]}   &                                        \\
\midrule
\multirow{2}{*}{17:15:00} & \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}    &  \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
\multirow{2}{*}{17:20:00}    & \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}} \\
         & \\[.25cm] 
17:21:00 & E123 captain and firefighter on second floor with 1\sfrac{3}{4}-inch hoseline. \\[.25cm]
17:21:30 & First floor exterior porch door opened. \\[.25cm]
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}} \\
         & \\[.25cm]
17:24:00    & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    & SQ5 makes entry through front of structure. \\[.25cm]
17:27:00    & After hearing a Mayday call from SQ5, IC calls a Mayday. \\[.25cm]
17:29:00    & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2} story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber with 12.7~mm (0.5~in) gypsum board over top~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}. The fire, which is described in Section~\ref{fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~\ref{fig:alpha_ex} shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the attic and rear porch on the second floor. Fully dimensioned drawings of the interior of the first and second floors are shown in Figs.~\ref{fig:first_floor} and \ref{fig:second_floor}, respectively.

@@

To estimate the type of fuel, fire size, and structure ventilation, on-scene videos and post-incident reports were used. Based on these findings, the FDS model includes two source fire locations: an initial fire in the attic and a secondary fire on the second floor porch landing. The attic and rear porch primarily consisted of wood fuel: the rear porch was enclosed with wood planks covered with foam board and vinyl siding, the roof consisted of asphalt shingles over wood planking, and the attic and porch floors were composed of oriented strand board.

@@

The value for the heat of combustion of wood used in this simulation was~16,400 kJ/kg, based on data provided in the SFPE Handbook~\cite{SFPE:Tewarson}. The heat of combustion quantifies the amount of energy per unit mass of the fuel. The following text defines Eq.~\ref{eq:wood_comb} for an FDS input file with the desired fuel and reaction properties discussed above:

@@

To estimate the energy release rate per unit area (kW/m$^2$) of wood, work by Babrauskas and Grayson~\cite{babrauskas1990} is used. Experiments were conducted in a cone calorimeter\footnote{The cone calorimeter is an experimental apparatus used to gather data about ignition time, mass loss, combustion products, and heat release rate among other properties associated with burning small samples of materials.} to determine the 5 minute average of HRRPUA\footnote{HRRPUA is the heat release rate per unit area.} (kW/m$^2$) for several different types of wood over a range of radiant heat fluxes (kW/m$^2$). Babrauskas and Grayson~\cite{babrauskas1990} suggest that the heat release rate per unit area for a variety of woods is approximately 50~kW/m$^2$.

To calculate the total fire size, the appropriate burning area (exposed wood area) of the attic and second floor porch enclosure needs to be determined. Based on the floor plan of the structure, the underside of the roof is approximately 134~m$^2$ (1453~ft$^2$) with 7~m$^2$ (75~ft$^2$) walls on the front and rear sides. Additionally, there is 105~m$^2$ (1130~ft$^2$) of floor area in the attic. Examination of post-incident images indicated that the floor immediately over the rear porch burned away. That area was considered to be part of the porch fire. Based on post-incident damage, approximately 50~\% of the remaining floor area and 90~\% of the roof was considered to be fuel for the attic fire. To estimate the total attic fire size, the underside roof area (120~m$^2$), two end walls (14~m$^2$), and the floor area (43~m$^2$ or 463~ft$^2$) were included. Applying an HRRPUA of wood of 50~kW/m$^2$ results in a potential peak fire size of 9~MW. To estimate the size of the porch fire, the four walls of the porch enclosure (42~m$^2$ or 452~ft$^2$) and the floor and ceiling of the porch (24~m$^2$ or 258 ft$^2$) were included, which results in a total porch area of 66~m$^2$ (689~ft$^2$). Using the same HRRPUA of wood of 50~kW/m$^2$, the potential peak fire size on the porch is approximately 3.3~MW. Therefore, the maximum fire size for this structure was estimated to be 12.3~MW. Since the time at which the fire started is not known relative to the notification of a fire or how fast the fire spread, several simplifications were made. The first assumption is that the attic fire was specified to start at the beginning of the simulation and increase to a value of 9~MW in 10~s. The porch fire was specified to start 40~s into the simulation and increase to a value of 3.3~MW in 10~s. The 10-s ramp up time was used to prevent abrupt changes in the velocity at the fire boundary condition in FDS. The second assumption is that the sources of the gas phase wood fuel have constant areas at fixed locations (2 sources for the attic, 1 source for the porch). The following lines define the source fires in the FDS input file:

@@

@@

An unknown parameter in Table~\ref{tab:vents} that needed to be estimated was the time when the second floor interior/porch door failed (2~min and 40~s into the simulation). The area of failure was also estimated based on the following observations. The bottom hinge from the door appeared to be closed throughout the fire due to the lack of fire damage and the lack of soot deposited on the interior faces of the hinge and remaining wood, as shown in Fig.~\ref{fig:door_frame}. It appears that the door collapsed and failed partially open. In the simulations, only the upper portion of the door was assumed to fail. While this time and area may not be exact, the takeaway from the analysis in the following sections should be impact of the door's failure on the interior conditions of the structure.

\begin{figure}[!ht]
\centering
\includegraphics[width=.80\textwidth]{../Figures/door_frame_burn}
\caption[Post-incident image of rear porch door frame.]{Post-incident image of the second floor porch door frame that failed taken from inside the structure looking toward the porch. Note that the hinge is still intact and there is unburned, ``clean'' wood at along the bottom portion of the frame.}
\label{fig:door_frame}
\end{figure}

\begin{table}
\centering
\captionof{table}{Timeline of ventilation changes in the simulation}\label{tab:vents}
\begin{tabular}{cccll}
\toprule[1.5pt]
Baseline    &  Alternative  &  Ventilation  &  Type         &  Side of         \\
Simulation  &  Simulation   &  Area         &               &  Structure       \\
{[s]}       &  {[s]}        &  [m$^2$]      &               &                  \\
\midrule
0           &  0            &  0.11         &  Window       &  Rear            \\
40          &  40           &  0.08         &  Breach (x2)  &  Front and Rear  \\
45          &  45           &  0.12         &  Breach       &  Left            \\
55          &  55           &  0.12         &  Breach       &  Rear            \\
130         &  N/A          &  1.41         &  Door         &  Rear            \\
135         &  N/A          &  0.2          &  Breach       &  Left            \\
160         &  160          &  0.80         &  Door         &  Interior        \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

For the simulation in this study, a measure of how well the flow field is resolved can be estimated by using the non-dimensional expression $D^*/\dx$. Here, $D^*$ is the characteristic fire diameter, $\dx$ is the nominal size of a mesh cell, and $\dQ$ is the total heat release rate of the fire:
\begin{equation}
D^* = \left(
     \frac{\dQ}{\rho_\infty \, c_p \, T_\infty \, \sqrt{g} }
     \right)^\frac{2}{5} 
\label{eq:mesh}
\end{equation}   
From the FDS User Guide~\cite{FDS_Users_Guide}, the characteristic fire diameter is related to the characteristic fire size via the
relation $Q^* = (D^*/D)^{5/2}$. Here, $D$ is the physical diameter of the fire as it is specified in the simulation. Following Eq.~(\ref{eq:mesh}) and using 10 cm grid cells, the ratio of the characteristic fire diameter to the cell size ($D^*/\dx$) is 15.5 for the 3.3~MW porch fire and 23 for the 9~MW attic fire. Based on validation work performed for the U.S. Nuclear Regulatory Commission, $D^*/\dx$ values ranged between 4 and 16~\cite{NUREG_1824} and produced results that were judged to be adequate for engineering calculations. The grid resolution used in this model is within typical engineering $D^*/\dx$ values. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas will snap to the nearest 10~cm. Figure~\ref{fig:geom_grid} shows the front of the structure rendered in Smokeview with the 10 cm mesh.

@@

As discussed in Section~\ref{geom}, the structure has dimensions of 17.15~m (56~ft) by 6.15~m (20~ft) and a total height of 8.56~m (28~ft). To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The total volume of the computational domain was 24~m (79~ft) by 10~m (33~ft) by 10~m (33~ft). At a grid resolution of 10~cm (3.9~in), this results in 2.4 million computational cells. The domain was divided into 16 equally sized meshes (each containing 150,000 grid cells) that could be processed in parallel, which reduced the amount of required calculation time to approximately 1.5~days. Figure~(\ref{fig:mult_mesh}) shows the entire structure within the computational domain that has been divided into multiple meshes. For the simulations that were conducted as part of this study, Table~\ref{tab:mod_param} shows the relevant model input parameters that were discussed in Section~\ref{model}.

@@

To examine the results of the simulations, it is important to link the timelines from fire scene to the simulation times. Table~\ref{tab:firesim} shows the fireground timeline along with the corresponding simulation times.

>>>>

% Extra packages
\usepackage{xfrac}
\usepackage{cite}

@@

\chapter*{\centering Abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on November 2, 2012, within a 2 \sfrac{1}{2} story, single family structure in Chicago, IL, that resulted in the death of the captain of a fire department. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at a height of 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

@@

On November 2, 2012, a fire in a 2~\sfrac{1}{2} story wood frame residential structure claimed the life of a captain of the Chicago Fire Department. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{fire_sum} provides a summary of the fire incident, Section~\ref{model} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{results} presents the simulation results, and Section~\ref{discussion} discusses the simulation results as they relate to firefighter safety and effectiveness.

\chapter{Summary of the Fire Incident}
\label{fire_sum}
The account of events for this incident was documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines a rapid intervention team (RIT) response to be one truck company, one battalion chief, one advanced life support (ALS) ambulance, and one emergency medical services (EMS) field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

@@

\noindent Table~\ref{tab:fire_info} shows an overview of the timeline of events. A plan view of the structure and the location where the captain collapsed are shown in Fig.~\ref{fig:simp_geom}, and more details of the fire development are provided in the following sections.

\begin{table}
\centering
\captionof{table}[Abridged NIOSH approximate fire event timeline]{Abridged NIOSH approximate fire event timeline~\cite{NIOSH:Bowyer}}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time              &  Fire Behavior / Fireground Operation                                                                                                    \\
{[hh:mm:ss]}               &                                                                                                                                          \\
\midrule
\multirow{2}{*}{17:15:00}  &  \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}}        \\ 
                           &                                                                                                                                          \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}  &   \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}}    \\
                           &                                                                                                                                          \\[.25cm]
\multirow{2}{*}{17:20:00}  &  \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}}      \\
                           &                                                                                                                                          \\[.25cm] 
17:21:00                   &  E123 captain and firefighter on second floor with 1\sfrac{3}{4}-inch hoseline.                                                          \\[.25cm]
17:21:30                   &  First floor exterior porch door opened.                                                                                                 \\[.25cm]
\multirow{2}{*}{17:23:00}  &  \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}}  \\
                           &                                                                                                                                          \\[.25cm]
17:24:00                   &   Water applied to attic from the rear exterior position.                                                                                \\[.25cm]
17:25:00                   &   SQ5 makes entry through front of structure.                                                                                            \\[.25cm]
17:27:00                   &   After hearing a Mayday call from SQ5, IC calls a Mayday.                                                                               \\[.25cm]
17:29:00                   &   E123 captain carried out of structure.                                                                                                 \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2} story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber covered by 12.7~mm (0.5~in) gypsum board~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}.

The fire, which is described in Section~\ref{fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~\ref{fig:alpha_ex} shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the attic and rear porch on the second floor. Fully dimensioned drawings of the interior of the first and second floors are shown in Figs.~\ref{fig:first_floor} and \ref{fig:second_floor}, respectively.

@@

To estimate the type of fuel, fire size, and structure ventilation, on-scene videos and post-incident reports were used. Based on these sources of information, the FDS model includes two source fire locations: an initial fire in the attic and a secondary fire on the second floor porch landing. The attic and rear porch primarily consisted of wood fuel: the rear porch was enclosed with wood planks covered with foam board and vinyl siding, the roof consisted of asphalt shingles over wood planking, and the attic and porch floors were composed of oriented strand board.

@@

The value for the heat of combustion of wood used in this simulation was~16,400 kJ/kg, based on data provided in the Society of Fire Protection Engineers (SFPE) Handbook~\cite{SFPE:Tewarson}. The heat of combustion quantifies the amount of energy per unit mass of the fuel. The following text defines Eq.~\ref{eq:wood_comb} for an FDS input file with the desired fuel and reaction properties discussed above:

@@

To estimate the heat release rate per unit area (HRRPUA) of wood, Babrauskas and Grayson~\cite{babrauskas1990} conducted experiments in a cone calorimeter\footnote{The cone calorimeter is an experimental apparatus used to gather data about ignition time, mass loss, combustion products, and heat release rate among other properties associated with burning small samples of materials~\cite{ASTM:E1355}.} to determine the 5-min average of the HRRPUA for several different types of wood over a range of radiant heat fluxes. The results of that study indicate that the HRRPUA for different types of wood is approximately 50~kW/m$^2$.

To calculate the total fire size, the appropriate burning area (exposed wood area) of the attic and second floor porch enclosure needs to be determined. Based on the floor plan of the structure, the underside of the roof was approximately 134~m$^2$ (1453~ft$^2$) with 7~m$^2$ (75~ft$^2$) walls on the front and rear sides. Additionally, there was 105~m$^2$ (1130~ft$^2$) of floor area in the attic. Examination of post-incident images indicated that the floor immediately over the rear porch burned away. That area was considered to be part of the porch fire. Based on post-incident damage, approximately 50~\% of the remaining floor area and 90~\% of the roof was considered to be fuel for the attic fire. To estimate the total attic fire size, the underside roof area (120~m$^2$), two end walls (14~m$^2$), and the floor area (43~m$^2$ or 463~ft$^2$) were included. Applying an HRRPUA of wood of 50~kW/m$^2$ results in a potential peak fire size of 9~MW. To estimate the size of the porch fire, the four walls of the porch enclosure (42~m$^2$ or 452~ft$^2$) and the floor and ceiling of the porch (24~m$^2$ or 258 ft$^2$) were included, which results in a total porch area of 66~m$^2$ (689~ft$^2$). Using the same HRRPUA of wood of 50~kW/m$^2$, the potential peak fire size on the porch was approximately 3.3~MW. Therefore, the maximum fire size for this structure was estimated to be 12.3~MW. Since the time at which the fire started is not known relative to the notification of a fire or how fast the fire spread, several simplifications were made. The first assumption is that the attic fire was specified to start at the beginning of the simulation and increase to a value of 9~MW in 10~s. The porch fire was specified to start 40~s into the simulation and increase to a value of 3.3~MW in 10~s. The 10-s ramp up time was used to prevent abrupt changes in the velocity at the fire boundary condition in FDS. The second assumption is that the sources of the gas phase wood fuel have constant areas at fixed locations (2 sources for the attic, 1 source for the porch). The following lines define the source fires in the FDS input file:

@@

\clearpage

@@

An unknown parameter in Table~\ref{tab:vents} that needed to be estimated was the time when the second floor interior/porch door failed (160~s into the simulation). The area of failure was also estimated based on the following observations. The bottom hinge from the door appeared to be closed throughout the fire due to the lack of fire damage and the lack of soot deposited on the interior faces of the hinge and remaining wood, as shown in Fig.~\ref{fig:door_frame}. It appears that the door collapsed and failed partially open. In the simulations, only the upper portion of the door was assumed to fail. While this time and area may not be exact, the takeaway from the analysis in the following sections should be impact of the door's failure on the interior conditions of the structure.

\begin{figure}[!ht]
\centering
\includegraphics[width=.80\textwidth]{../Figures/door_frame_burn}
\caption[Post-incident image of rear porch door frame.]{Post-incident image of the second floor porch door frame that failed taken from inside the structure looking toward the porch. Note that the hinge is still intact and there is unburned, ``clean'' wood along the bottom portion of the frame.}
\label{fig:door_frame}
\end{figure}

\begin{table}
\centering
\captionof{table}{Timeline of ventilation changes in the simulation}\label{tab:vents}
\begin{tabular}{cccll}
\toprule[1.5pt]
Baseline    &  Alternative  &  Ventilation  &  Type         &  Side of         \\
Simulation  &  Simulation   &  Area         &               &  Structure       \\
{[s]}       &  {[s]}        &  [m$^2$]      &               &                  \\
\midrule
0           &  0            &  0.11         &  Window       &  Rear            \\
40          &  40           &  0.08         &  Breach (x2)  &  Front and Rear  \\
45          &  45           &  0.12         &  Breach       &  Left            \\
55          &  55           &  0.12         &  Breach       &  Rear            \\
130         &  N/A$^*$      &  1.41         &  Door         &  Rear            \\
135         &  N/A$^*$      &  0.2          &  Breach       &  Left            \\
160         &  160          &  0.80         &  Door         &  Interior        \\
\bottomrule[1.25pt]
\end{tabular}\par
\footnotesize
$^*$These ventilation events were not included in the alternative simulation.
\normalsize
\end{table}

@@

For the simulation, a measure of how well the flow field is resolved can be estimated by using the non-dimensional expression $D^*/\dx$. Here, $D^*$ is the characteristic fire diameter, $\dx$ is the nominal size of a mesh cell, and $\dQ$ is the total heat release rate of the fire:
\begin{equation}
D^* = \left(
     \frac{\dQ}{\rho_\infty \, c_p \, T_\infty \, \sqrt{g}}
     \right)^\frac{2}{5} 
\label{eq:mesh}
\end{equation}   
From the FDS User's Guide~\cite{FDS_Users_Guide}, the characteristic fire diameter is related to the characteristic fire size via the
relation $Q^* = (D^*/D)^{5/2}$. Here, $D$ is the physical diameter of the base of the fire specified in the simulation. Following Eq.~(\ref{eq:mesh}) and using a grid cell size of 10~cm, the characteristic fire diameter to cell size ($D^*/\dx$) ratio is 15.5 for the 3.3~MW porch fire and 23 for the 9~MW attic fire. Based on validation work performed for the U.S. Nuclear Regulatory Commission, $D^*/\dx$ values ranged between 4 and 16~\cite{NUREG_1824} and produced results that were judged to be adequate for engineering calculations. The grid resolution used in this model is within typical engineering $D^*/\dx$ values. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas will snap to the nearest 10~cm. Figure~\ref{fig:geom_grid} shows the front of the structure rendered in Smokeview with the 10 cm mesh.

@@

As discussed in Section~\ref{geom}, the structure has dimensions of 17.15~m (56~ft) by 6.15~m (20~ft) and a total height of 8.56~m (28~ft). To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The total volume of the computational domain was 24~m (79~ft) by 10~m (33~ft) by 10~m (33~ft). A grid resolution of 10~cm (3.9~in) was used, which required 2.4 million computational cells. The domain was divided into 16 equally sized meshes (each containing 150,000 grid cells) that could be processed in parallel, which reduced the amount of required calculation time to approximately 1.5~days. Figure~(\ref{fig:mult_mesh}) shows the entire structure within the computational domain that has been divided into multiple meshes. For the simulations that were conducted as part of this study, Table~\ref{tab:mod_param} shows the relevant model input parameters that were discussed in Section~\ref{model}.

@@

To examine the results of the simulations, it is important to link the timelines from the fire scene to the simulation times. Table~\ref{tab:firesim} shows the fireground timeline~\cite{NIOSH:Bowyer} along with the corresponding simulation times.
